# Continuous Probability Distributions

## Learning Objectives {-#objectives-continuous-distributions}

1. Define and explain the key characteristics of the continuous distributions:
  - normal, 
  - lognormal, 
  - exponential, 
  - gamma, 
  - chi-square, 
  - $t$, 
  - $F$, 
  - beta and 
  - uniform on an interval.
2. Evaluate probabilities and quantiles associated with such distributions.
3. Generate discrete and continuous random variables using the inverse transform method.

## Theory {-#theory-continuous-distributions}

**A reminder:** `R` was designed to be used for statistical computing - so it handles **randomness** well! Using `R` we can guarantee reproducibility (and enhance sharability) by using the function `set.seed(seed)`{.R} where `seed` is a single value integer. Using this approach we guarantee the generation of the same sequence of random numbers everytime we call this function. Use `?set.seed`{.R} to learn more about this function.

## In-built probability distributions

**A recap:** `R` has in-built functions for probability distributions:

- **d***\<distribution-name\>* $:=$ **d**ensity ("_PDF_"), *i.e.* $f_X(x)$
- **p***\<distribution-name\>* $:=$ **p**robability distribution cumulative function ("_CDF_"), *i.e.* $F_X(x) =\boldsymbol{P}(X \leq x)$
- **q***\<distribution-name\>* $:=$ **q**uantile function, *i.e.* return $x$ such that $\boldsymbol{P}(X \leq x) = p$
- **r***\<distribution-name\>* $:=$ **r**andom deviates, *i.e.* (*psuedo*) random number generator for a given distribution
- *Where \<distribution-name\>* $=$ Normal, uniform, lognormal, Student's $t$, Poisson, binormal, Weibull ... see `?distributions()`{.R} for more information

To give some quick examples (we will further explore these in more detail later in this chapter):

| R Code | Definition |
|-|-|
| `rnorm(1)`{.R} | Generates $x_1$ where $X \sim \mathcal{N}(0,\,1)$ |
| `rnorm(y, mean=10, sd=2)`{.R} | Generates $\{y_1,\,y_2,\,\dots\}$ with $Y \sim \mathcal{N}(10,\,2^2)$ |
| `runif(3, min=5, max=10)`{.R} | Generates $\{z_1,\,z_2,\,z_3\}$  where $Z \sim \mathcal{U}(5,\,10)$ |
| `dbinom(4, size=5, prob=0.5)`{.R} | Computes $\boldsymbol{P}(X = 4)$ where $X \sim Bin(5,\,0.5)$ |
| `pgamma(0.2, shape=2, rate=2)`{.R} | Computes $F_Y(0.2)$ where $Y \sim \mathcal{\Gamma}(2,\,2)$, i.e.  $\boldsymbol{P}(Y\leq 0.2)$|
| `qexp(0.5, rate = 2)`{.R} | Determines smallest value of $z$ for $\boldsymbol{P}(Z \leq z) = 0.5$ where $Z \sim Exp(2)$ |

## Continuous probability distributions covered {-}

We will consider how to interact with the following continuous probability distributions in `R`:

- Normal
- Lognormal
- Exponential
- Gamma
- $\chi^2$
- Student's $t$
- $F$
- Beta
- Uniform (on an interval)

For each distribution above we will determine how to calculate:

- A random deviate following the discrete distribution $X$,
- The probability density function ("_PDF_"), $P(k_1 \leq X \leq k_2)$ for distribution $X$ over the range $[k_1,\,k_2]$,
- The cumulative distribution function ("_CDF_"), $P(X \leq k)$, and
- The quantile function to find $k$ representing the value such that $P(X \leq k) = p$, i.e. the _pth_ percentile.

We will finish off with a plot of the distribution.

## Normal distribution

We start with generating random deviates from the Normal distribution.

If we are interested in the standard normal, `R` helpfully has default argument values for $\mu = 0$ and $\sigma = 0$ so the function call is very concise:

```{r continuous-distributions-normal-1}
# Generate random deviates
set.seed(42)
rnorm(5)
```

We can also specify our own values of $\mu$ and $\sigma$. With `R` we need to remember that the $\sigma$ argument corresponds to the standard deviation, **not** the variance.

```{r continuous-distributions-normal-2}
# Generate random deviates
set.seed(42)
rnorm(5, 10, 2)
```

We next look at the cumulative distribution function for a Normal distribution. In `R` we can calculate this using `pnorm(q, mean, sd)`{.R} where:

- _q_ is the quantile of interest,
- _mean_ is the mean, and
- _sd_ is standard deviation.

```{r continuous-distributions-normal-3}
# Calculate P(X <= 2) for X~N(0,1)
```

Next we want to find the x<sup>th</sup> percentile of $X \sim N(\mu, \sigma)$. We use the quantile function, `qnorm(mu, sigma)`{.R}.

```{r continuous-distributions-normal-4}
# Find the 99th percentile for X~N(10,2)
percentile_99 <- qnorm(0.99, 10, 2)

paste0(
  "The 99th percentile of X~N(10, 2) is ",
  format(percentile_99, digits = 4),
  "."
)
```

As is customary we finish with a plot of the normal distribution.

```{r continuous-distributions-normal-5}
# Plot the distribution function of X~Normal(mu =, sigma = )
```

## Lognormal distribution

```{r continuous-distributions-lognormal-1}

# Plot the distribution function of Y~Lognormal()
```

## Exponential distribution

```{r continuous-distributions-exponential-1}

# Plot the distribution function of Z~Exp()
```

## Gamma distribution

```{r continuous-distributions-gamma-1}

# Plot the distribution of X~Gamma()
```

## $\chi^2$ distribution

```{r continuous-distributions-chi-squared-1}

# Plot the distribution of Y~Chi-Square()
```

## Student's $t$ distribution

```{r continuous-distributions-t-1}

# Plot the distribution of Z~t()
```

## $F$ distribution

```{r continuous-distributions-F-1}

# Plot the distribution of X~F()
```

## Beta distribution

```{r continuous-distributions-Beta-1}

# Plot the distribution of Y~Beta()
```

## Uniform distribution {#continuous-uniform}

```{r continuous-distributions-uniform-1}

# Plot the distribution of Z~Uniform()
```

## Inverse transform method

The inverse transform method is a way to generate psuedo-random numbers from any probability distribution. 

One possible algorithm is as follows:

1. Generate a random number $u$ from $U \sim \mathcal{U}(0, 1)
2. Find the inverse of the desired cumulative distribution function, $F^{-1}_X(x)$
3. Compute $X = F^{-1}_X(u)$

Suppose we wanted to draw 10,000 random numbers from $X \sim Exp(\lambda = 2)$. In order to use the inverse transform method we first need to find the inverse of the CDF. For any $X \sim Exp(\lambda)$, the inverse of the CDF is $\frac{-log(1-x)}{\lambda}$. We can thus use the inverse transform algorithm to generate random deviates following $X \sim Exp(2)$:

```{r continuous-distributions-inverse-transform-1}
# Step 0 - to guarantee reproducibility
set.seed(42)

# Step 1 - generate 10,000 random deviates from U[0,1]
u <- runif(10000)

# Step 2 - find the inverse of the CDF: 1 - exp(-lambda.x)
# Inverse of CDF = -log(1 - x) / lambda

# Step 3 - compute X using the inverse of the CDF [from step 2] and the random deviates u [from step 1]
x <- -log(1 - u) / 2

# Plot the resulting x deviates
library(ggplot2)
df <- data.frame(x = x) 
ggplot(df, aes(x=x)) +
  geom_histogram(binwidth = 0.5, colour="black", fill="white")
```

## `R` Practice {-#practice-continuous-distributions}

We finish with a comprehensive example of an univariate continuous distribution question in `R`.